---
title: "Malaria Data Analysis Report"
subtitle: "A Comparative Analysis of Epidemiological Patterns"
author: "Workshop Participant"
date: "`r Sys.Date()`"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 10
    fig-height: 6
    dpi: 300
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    toc: true
    number-sections: true
    fig-width: 6
    fig-height: 4
    dpi: 300
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required libraries
library(tidyverse)
library(ggplot2)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
```

# Abstract

This report presents a comprehensive analysis of malaria surveillance data from the Epi R Handbook, examining patterns in case counts by age group, district-level variations, and reporting delays over the period May-August 2020. The analysis employs both traditional statistical methods and modern data visualization techniques to identify key epidemiological trends and inform malaria control programme decision-making.

# Introduction

Malaria remains a significant public health challenge in many regions worldwide. Understanding the temporal patterns of disease transmission, age-specific incidence, and surveillance system performance is crucial for developing effective control strategies. This analysis examines malaria surveillance data from four districts to identify common patterns and district-specific variations in disease burden and reporting quality.

## Objectives

The primary objectives of this analysis are to:

1. Characterise the temporal patterns of malaria cases by age group and district
2. Compare disease burden across districts using standardised metrics
3. Examine age-specific incidence patterns over time
4. Assess surveillance system performance through reporting delay analysis

# Methods

## Data Source

The analysis utilises malaria surveillance data from the Epi R Handbook representing facility-level case counts from four districts over the period 16 May 2020 to 12 August 2020. The dataset includes:

- Daily case counts by age group (0-4, 5-14, 15+ years)
- Facility-level data from 65 health facilities
- District identifiers (Spring, Bolo, Dingo, Barnard)
- Data collection and submission dates
- Facility identifiers

## Data Processing

Data processing involved several key steps:

1. **Data cleaning**: Handling missing values and data quality assessment
2. **Age group analysis**: Calculation of age-specific proportions and totals
3. **Temporal aggregation**: Creation of weekly summaries by district
4. **Reporting delay calculation**: Computation of days between data collection and submission

## Statistical Analysis

Statistical analyses included:

- Descriptive statistics by district and time period
- Trend analysis describing temporal patterns in reported cases
- Peak period identification and comparison
- Age-specific burden assessment across districts

## Software and Tools

Analysis was conducted using R version 4.3.0 with the following packages:
- `tidyverse` for data manipulation
- `ggplot2` for visualization
- `lubridate` for date handling
- `scales` for formatting

# Results

## Data Overview

```{r data-overview}
# Load and examine the dataset
# Note: In the actual workshop, participants will load their processed data
# This is a placeholder showing the expected structure

# Create sample data structure for demonstration
sample_data <- data.frame(
  district = c("Spring", "Bolo", "Dingo", "Barnard"),
  total_cases = c(34333, 32195, 24556, 13586),
  total_0_4 = c(12351, 9411, 7546, 4579),
  total_5_14 = c(7954, 6590, 6380, 3700),
  total_15_plus = c(14028, 16194, 10630, 5307),
  num_facilities = c(65, 65, 65, 65)
)

kable(sample_data, 
      caption = "Summary of malaria cases by district and age group",
      col.names = c("District", "Total Cases", "0-4 years", "5-14 years", "15+ years", "Facilities")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Temporal Patterns

### Daily Case Trends

```{r daily-cases, fig.cap="Weekly malaria cases by district, May-August 2020"}
# Placeholder for actual plot code
# Participants will generate this plot during the workshop

# Create sample plot structure
ggplot() +
  labs(
    title = "Weekly Malaria Cases by District",
    subtitle = "May-August 2020",
    x = "Date",
    y = "Number of Cases",
    caption = "Data source: Epi R Handbook Malaria Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

### Age Distribution

```{r standardised-rates, fig.cap="Age group distribution of malaria cases over time"}
# Placeholder for age group distribution plot
# This will show age group proportions over time

ggplot() +
  labs(
    title = "Age Group Distribution of Malaria Cases",
    subtitle = "Proportion of cases by age group and district",
    x = "Date",
    y = "Proportion of Cases",
    caption = "Data source: Epi R Handbook Malaria Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

## Reporting Delays

```{r case-fatality, fig.cap="Reporting delay analysis by district"}
# Placeholder for reporting delay plot
# This will show surveillance system performance

ggplot() +
  labs(
    title = "Average Reporting Delay by District",
    subtitle = "Days between data collection and submission",
    x = "Date",
    y = "Average Reporting Delay (Days)",
    caption = "Data source: Epi R Handbook Malaria Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )
```

## Peak Period Analysis

```{r peak-analysis}
# Placeholder for peak period analysis
# This will identify and compare peak transmission periods

peak_summary <- data.frame(
  district = c("Spring", "Bolo", "Dingo", "Barnard"),
  peak_week_cases = c(6380, 5605, 4522, 3300),
  peak_week_0_4 = c(2249, 1502, 1299, 1200),
  peak_week_5_14 = c(1575, 1022, 1250, 927),
  peak_week_15_plus = c(2556, 3081, 1973, 1173)
)

kable(peak_summary,
      caption = "Peak week characteristics by district",
      col.names = c("District", "Peak Total Cases", "Peak 0-4 years", "Peak 5-14 years", "Peak 15+ years")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Discussion

## Key Findings

Use this section to summarise the most important insights from your analysis. Highlight temporal trends, district-level contrasts, age-specific burdens, and reporting performance.

## Implications for Public Health

Discuss how the evidence could inform surveillance improvements, allocation of vector-control resources, or communication with district teams.

## Limitations

Outline data quality caveats, representativeness concerns, and any analytical shortcuts taken during the workshop.

## Future Directions

Propose ideas for deeper analysis (e.g., facility-level exploration, integration with climate indicators, longer time series) that participants could pursue after the session.

# Conclusion

Reinforce the main takeaways from the workshop exercise and connect them to participants’ own malaria surveillance work.

# References

1. World Health Organization. World Malaria Report 2020. Geneva: WHO; 2020.
2. Applied Epi. The Epidemiologist R Handbook. Available at: https://www.epirhandbook.com/
3. R Core Team. R: A language and environment for statistical computing. Vienna: R Foundation for Statistical Computing; 2021.

# Appendix

## R Session Information

```{r session-info}
sessionInfo()
```

## Code Repository

The complete analysis code is available in the workshop materials, demonstrating three different approaches to data analysis:

1. **Manual Coding**: Traditional R programming approach
2. **Copilot-Assisted**: AI-enhanced coding with GitHub Copilot
3. **AI Agent**: Advanced natural language programming with VS Code

---

*This report was generated as part of the AI-Augmented R Coding workshop, demonstrating the integration of artificial intelligence tools in epidemiological data analysis.*
